\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
cs1 redone

options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(ggplot2)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv) 
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(visreg)
library(lubridate)
library(tidyverse)
library(scales)
options(digits = 2)
# set the seed for reproducible output
set.seed(1234)

@@

```{r erspc-data, eval = eval_cs1}
pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa", exposure = "ScrArm")
inherits(pt_object, "popTime")
attr(pt_object, "exposure")
```

@@

For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. We fit a flexible hazard by using a smooth function of time modeled with cubic B-splines. The model is fit using \code{fitSmoothHazard} with the familiar formula interface:

```{r, eval = eval_cs1, echo = -1}
set.seed(12)
library(splines)
fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                       data = ERSPC, ratio = 100)
```

The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes. For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:

@@

@@

```{r, eval=eval_cs1, echo=echo_plot_code}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

@@

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate. As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the significance of the spline model:

```{r erscp-compare, eval = eval_cs1, echo = FALSE}
anova(fit, test = "LRT")
```

The null hypothesis is rejected in favor of the spline model. Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.

@@

```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

@@

Next, we can compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.

```{r eval = eval_cs1, echo = -1}
new_time <- c(0, 1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
conf_ints <- confint(risk, fit)
conf_ints
```

These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates.

```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
# confidence interval
tt <- confint(risk_mp, fit)

# plot absrisk and 95% CI
new_time <- seq(0,14,0.1)
i.backw <- order(new_time, decreasing = TRUE)
i.forw <- order(new_time)

@@

Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality]. The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed. As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward. This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model:

```{r, echo=-1, eval = eval_cs1}
set.seed(12)
fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
                             data = ERSPC)
```

Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package. In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard ratio and 95% confidence interval for screening vs. control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:

```{r, echo = TRUE, eval = FALSE}
plot(fit_inter, type = "hr", newdata = new_data,
     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
```

```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm. 95\\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)

par(oma = c(0,1,0,8))
tt <- plot(fit_inter,
     type = "hr",
     newdata = new_data,
     var = "ScrArm",
     increment = 1,
     xvar = "Follow.Up.Time",
     ci = T,
     xlab = "Follow-up time (years)",
     ylab = "Death from prostate cancer hazard ratio",
     ylim = c(0,1.50),
     yaxt = 'n',
     xaxt = "n",
     rug = TRUE)

@@

We want to compare our hazard ratio estimates to that obtained from a Cox regression.

@@

```{r supportData, eval = eval_cs3}
scaleFUNy <- function(x) sprintf("%.2f", x)
scaleFUNx <- function(x) sprintf("%.0f", x)
data(support)
# Change time to years
support$d.time <- support$d.time/365.25

>>>>

options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(survival)
library(splines)
library(ggplot2)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv)
library(prodlim)
# library(riskRegression)
library(tibble)
library(visreg)
library(lubridate)
library(tidyverse)
library(scales)
options(digits = 2)
# set the seed for reproducible output
set.seed(1234)

@@

```{r erspc-data, eval = eval_cs1, echo=TRUE}
pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa", exposure = "ScrArm")
inherits(pt_object, "popTime")
attr(pt_object, "exposure")
```

@@

For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. 
Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality]. 
The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed. 
As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward. 
This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model. 
We fit a flexible hazard by using a smooth function of time modeled with a penalized cubic spline basis with 2 degrees of freedom (implemented in the \code{survival::pspline} function). 
The model is fit using \code{fitSmoothHazard} with the familiar formula interface:

```{r, eval = eval_cs1, echo = -1}
set.seed(1953)
fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                       data = ERSPC, ratio = 100)
```

Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package. 
The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:

@@

As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. 
Therefore, we can easily test the significance of the spline term and its interaction with time:

```{r erscp-compare, eval = eval_cs1, echo = TRUE}
anova(fit, test = "LRT")
```

Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.

### Time-dependent hazard ratios

In what follows, the hazard ratio for a variable $X$ is defined as

$$
\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
$$
where $\lambda(t|\cdot;\hat{\beta})$ is the hazard rate as a function of the variable $t$ (which is usually time, but can be any other continuous variable), $x_1$ is the value of $X$ for the exposed group, $x_0$ is the value of $X$ for the unexposed group, $\mathbf{Z}$ are other covariates in the model which are equal to $\mathbf{z_1}$ in the exposed and $\mathbf{z_0}$ in the unexposed group, and $\hat{\beta}$ are the estimated regression coefficients. 
As indicated by the formula above, it is most instructive to plot the hazard ratio as a function of a variable $t$ only if there is an interaction between $t$ and $X$. 
Otherwise, the resulting plot will simply be a horizontal line across time.

The \code{plot} method with \code{type="hr"} for objects of class \code{singleEventCB} can be used to compute time-dependent hazard ratios and confidence intervals. 
In Figure \ref{fig:interaction-ERSPC}, we show the estimated hazard ratio and 95% confidence interval for screening vs. control group as a function of time. 
Note that we must specify the covariate profiles and times for the predicted hazards.

```{r, echo = TRUE, eval = FALSE}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)
plot(fit, type = "hr", newdata = new_data,
     var = "ScrArm", xvar = "Follow.Up.Time", ci = T)
```

```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm. 95\\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)

par(oma = c(0,1,0,8))
tt <- plot(fit,
     type = "hr",
     newdata = new_data,
     var = "ScrArm",
     increment = 1,
     xvar = "Follow.Up.Time",
     ci = T,
     xlab = "Follow-up time (years)",
     ylab = "Death from prostate cancer hazard ratio",
     ylim = c(0,1.50),
     yaxt = 'n',
     xaxt = "n",
     rug = TRUE)

### Hazard functions

Modeling the hazard function directly allows us to easily visualize it with the \code{plot} method and \code{type="hazard"} for objects of class \code{singleEventCB}. 
We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}. 
The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data} C, where more events are observed at later follow-up times. 
The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.

```{r cs1hazard, eval=eval_cs1, echo=TRUE}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
```

### Absolute risk

@@

```{r, eval=FALSE, echo=FALSE}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

@@

```{r, eval=eval_cs1, echo=TRUE}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
head(risk)
```

<!--In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate. -->

@@

```{r erspc-cox-cif, eval = FALSE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

@@

We can subsequently compute confidence intervals for the risk function using the method \code{confint.absRiskCB}:

```{r eval = eval_cs1, echo = TRUE}
conf_ints <- confint(risk, fit)
head(conf_ints)
```

These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE. 
In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates. 
We also overlay the Kaplan-Meier curves as a reference.

```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}

tt <- conf_ints

# plot absrisk and 95% CI
i.backw <- order(new_time, decreasing = TRUE)
i.forw <- order(new_time)

@@

```{r}
knitr::knit_exit()
```

@@

We want to compare our hazard ratio estimates to that obtained from a Cox regression (using the \pkg{survival} package version 3.2-13).

@@

```{r supportData, eval = eval_cs3}
library(riskRegression)
scaleFUNy <- function(x) sprintf("%.2f", x)
scaleFUNx <- function(x) sprintf("%.0f", x)
data(support)
# Change time to years
support$d.time <- support$d.time/365.25
